---
title: "Spatial data in R: an introduction to the sf package"
author: "Henry Partridge"
date: "2017-06-05"
output: 
  html_notebook:
    code_folding: hide
    fig_caption: no
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
## Global code options
knitr::opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
```

***

### INSTALLATION AND LOADING

To install [sf](https://cran.rstudio.com/web/packages/sf/) from CRAN:
```{r eval=FALSE}
install.packages("sf")  
```

or the development version from GitHub:
```{r eval=FALSE}
devtools::install_github("edzer/sfr")
```

then load:
```{r}
library(sf)
```

**NB** MacOSX and LINUX users need to install a number of geospatial libraries (GEOS, GDAL, and proj.4).     

The [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html) package also needs to be loaded:
```{r}
library(tidyverse)
```

### EXAMPLE DATA     
A GeoJSON of Greater Manchester's wards was created from a vector boundary file available from [ONS's Open Geography Portal](http://geoportal.statistics.gov.uk/datasets/afcc88affe5f450e9c03970b237a7999_2). The GeoJSON is projected in British National Grid (EPSG:27700) and originally derives from the [Ordnance Survey](https://www.ordnancesurvey.co.uk/business-and-government/products/opendata-products.html).

Point data are supplied by [data.police.uk](http://data.police.uk) and represent incidents of anti-social behaviour and crime recorded by the Greater Manchester Police during February 2017. The incidents are supplied with latitude and longitude coordinates which have undergone an [anonymisation process](http://data.police.uk/about/#anonymisation).

### READING AND WRITING SPATIAL DATA

Reading spatial data (polygons)
```{r}
bdy <- st_read("data/wards.geojson", quiet = TRUE)
```

Reading data with coordinates (points)
```{r}
pts <- read_csv("data/2017-03-greater-manchester-street.csv")
pts <- st_as_sf(pts, coords = c("Longitude", "Latitude"))
```

Writing spatial data
```{r eval=FALSE}
st_write(bdy, "boundaries.shp")
```

### SF OBJECTS

Attribute table (dataframe) AND geometry (list-column) with coordinates, CRS, bbox
```{r}
class(bdy)
head(bdy)
as.tibble(bdy)
bdy_df <- st_set_geometry(bdy, NULL) # remove geometry
head(bdy_df)
st_geometry(bdy) # print geometry
```

### MANIPULATE USING DPLYR

Rename variables
```{r}
bdy <- bdy %>% 
  select(ward = wd16nm, census_code = wd16cd, borough = lad16nm) # rename variables
glimpse(bdy)
```

Count frequency of wards by borough
```{r}
bdy %>% 
  group_by(borough) %>% 
  count() %>% 
  arrange(desc(n)) %>% # sort in descending order
  st_set_geometry(., NULL) # hide geometry
```

Select features
```{r}
chorlton <- bdy %>% 
  filter(ward == "Chorlton")
plot(st_geometry(bdy))
plot(st_geometry(chorlton), col = "red", add = TRUE)
```

Using sf functions to add geometry column to dplyr chain
```{r}
bdy <- bdy %>% 
  mutate(area = st_area(.)) # returns the area of a feature

bdy %>% 
  select(ward, area) %>% 
  arrange(desc(area)) %>% 
  slice(1:10) %>% 
  st_set_geometry(., NULL)
```

### PROJECTION

Check and assign CRS
```{r}
st_crs(pts) 
pts <- st_set_crs(pts, 4326) # assign Lat/Long (epsg:4326)
st_crs(pts)
```

Reproject CRS
```{r}
bdy_WGS84 <- st_transform(bdy, 4326)
st_crs(bdy_WGS84)
```

### CONVERT TO AND FROM SP OBJECTS

Convert to and from sp objects
```{r}
bdy_sp <- as(bdy, 'Spatial')
class(bdy_sp)
```

```{r}
bdy_sf <- st_as_sf(bdy_sp)
class(bdy_sf)
```

### SPATIAL OPERATIONS

Buffer features
```{r}
buffer <- chorlton %>% 
  st_buffer(dist = 1000)
plot(st_geometry(buffer))
plot(st_geometry(chorlton), col = "red", add = TRUE)
```

Buffer and intersect
```{r}
pts_sub <- bdy %>%
  filter(ward == "Chorlton") %>%
  st_buffer(dist = 1000) %>%
  st_intersection(st_transform(pts, 27700)) # reproject pts to BNG (epsg:27700)

plot(st_geometry(buffer))
plot(st_geometry(chorlton), col = "red", add = TRUE)
plot(st_geometry(pts_sub), col = "black", add = TRUE)
```

```{r}
pts_sub %>% 
  group_by(`Crime type`) %>%
  count() %>% 
  arrange(desc(n)) %>% 
  st_set_geometry(., NULL)
```

Points in polygon
```{r}
pts %>% 
  filter(`Crime type` == "Vehicle crime") %>%  
  st_join(bdy_WGS84, ., left = FALSE) %>% 
  count(ward) %>% 
  arrange(desc(n)) %>% 
  st_set_geometry(., NULL)

bdy_pts <- pts %>% 
  filter(`Crime type` == "Vehicle crime") %>%  
  st_join(bdy_WGS84, ., left = FALSE) %>% 
  count(ward)
```

### PLOTTING

Base plots
```{r}
plot(bdy_pts) # plots small multiples if dataframe has several attributes
```

```{r}
plot(bdy_pts["n"]) # select the appropriate attribute to plot a single map
```

```{r}
library(RColorBrewer) ; library(classInt)
pal <- brewer.pal(5, "RdPu")
classes <- classIntervals(bdy_pts$n, n=5, style="pretty")$brks
plot(bdy_pts["n"], 
     col = pal[findInterval(bdy_pts$n, classes, all.inside=TRUE)], 
     main = "Vehicle crime in Greater Manchester\nMarch 2017", axes = F)
legend("bottomright", legend = paste("<", round(classes[-1])), fill = pal, cex = 0.7) 
```

[ggplot2](https://cran.r-project.org/web/packages/ggplot2/index.html)
```{r, eval=FALSE}
devtools::install_github("tidyverse/ggplot2") # NB need development version for geom_sf()
```

```{r}
ggplot(bdy_pts) +
  geom_sf(aes(fill = n)) +
  scale_fill_gradientn('Frequency', colours=RColorBrewer::brewer.pal(5,"RdPu"), 
                       breaks = scales::pretty_breaks(n = 5)) +
  labs(fill = "Frequency",
       title = "Vehicle crime",
       subtitle = "March 2017",
       caption = "Contains OS data © Crown copyright and database right (2017)") +
  theme_void()
```

[plotly](https://cran.r-project.org/web/packages/plotly/index.html)
```{r}
library(plotly)
p <- ggplot(bdy_pts) +
  geom_sf(aes(fill = n, text = paste0(ward, "\n", "Crimes: ", n))) +
  scale_fill_gradientn('Frequency', colours=RColorBrewer::brewer.pal(5,"RdPu"), 
                       breaks = scales::pretty_breaks(n = 5)) +
  labs(fill = "Frequency",
       title = "Vehicle crime",
       subtitle = "March 2017",
       caption = "Contains OS data © Crown copyright and database right (2017)") +
  theme_void() + 
  coord_fixed(1.3)
ggplotly(p, tooltip = "text")
```

[leaflet](https://cran.r-project.org/web/packages/leaflet/index.html)
```{r}
library(leaflet)
pal <- colorBin("RdPu", domain = bdy_pts$n, bins = 5, pretty = TRUE)
leaflet(bdy_pts) %>% 
  addTiles(urlTemplate = "http://{s}.tiles.wmflabs.org/bw-mapnik/{z}/{x}/{y}.png",
    attribution = '&copy; <a href="http://www.openstreetmap.org/copyright">OpenStreetMap</a>, <a href="https://www.ons.gov.uk/methodology/geography/licences">Contains OS data © Crown copyright and database right (2017)</a>') %>% 
  addPolygons(fillColor = ~pal(n), fillOpacity = 0.8,
              weight = 1, opacity = 1, color = "black",
              label = ~as.character(ward)) %>% 
  addLegend(pal = pal, values = ~n, opacity = 0.7, 
            title = 'Vehicle crime (March 2017)', position = "bottomleft") %>%
  addMiniMap(tiles = providers$CartoDB.Positron, toggleDisplay = TRUE)
```